Outline of Ideas for BTC Transaction Frequency

### 2015-10-12 outline/outline.tex
Major update, still in progress. Flows much better and includes results from gamma fit showing a nonconstant rate is necessary

\section{The Basic Model}
Bitcoin blocks are by design distributed according a Poisson process with a rate of discovery of one block every 10 minutes. Let $\tau$ index time in minutes, and $N(\tau)$ denote the number of blocks discovered by time $\tau$. Then for $\delta > 0$, the number of blocks discovered between time $\tau$ and time $\tau + \delta$ is Poisson distributed, that is $N(\tau + \delta) - N(\tau) \sim Poi(\delta/10)$. Let $k=1,2,\dots$ denote each of the blocks, in order. Then the amount of time between blocks is exponentially distributed. Let $\tau_k$ denote the the time when block $k$ arrives. Then $\tau_{k+1} - \tau_{k} \stackrel{iid}{\sim} exp(1/10)$ for $k=1,2,\dots,K$. This part of the model is determined for us, but determining the how the number of transactions is distributed or, more precisely, how the amount of transaction {\it data} is distributed requires more work.

Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node, with a maximum offset of 70 minutes. Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block is completed. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and $t_k$ the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the chain at time $\tau$ so that $x_{\tau_{k+1}} - x_{\tau_k}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some stochastic process $dx_\tau$ so that $x_{\tau + \delta} - x_\tau = \int_\tau^{\tau + \delta}dx_\tau$, a stochastic integral. Then the amount of transaction data that arrives in $k$'th block is given by $\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau$.

Most of the interesting modeling choices come from modeling $dx_\tau$. Conventional Brownian motion is inappropriate since the increments should be nonnegative. On the other hand, geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider the classes of non-negative infinite activity L\'{e}vy processes with closed form solutions for $x_{\tau}$. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A better example for our purposes is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$. We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(alpha,\beta)$, it has the density
\[
p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
\]
for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$.

While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x\sim G(\alpha_x,\beta)$ independent of $y\sim G(\alpha_y,\beta)$, then $x+y \sim G(\alpha_x + \alpha_y, \beta)$. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error. This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_k + \delta_k}$ without resorting to approximation. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, but many reasonable models on these parameters will not allow for closed form distributions of the parameter increments.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes computational problems since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar model that avoids this problem is the lognormal distribution.

To start, consider the following parameterized model using the structure discussed above. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
\delta_{k} = \tau_{k} - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ --- the block has $\delta_k$ minutes to accumulate transaction data.

\subsection{Stochastic transaction rates --- seasonality and autocorrelation}
It is unlikely that the transaction rate is constant in time. The transaction rate is clearly growing as Bitcoin becomes more popular, and seasonal variation is expected throughout the data. Further, autocorrelation in transaction rates is likely since periods of high activity will sometimes be caused by, e.g., new information about the long term prospects of Bitcoin.

Let $\gamma_\tau$ denote the transaction rate at time $\tau$. We model the evolution of this rate as a variant of the exponential Ornstein–Uhlenbeck (OU) process. Specifically let $\varepsilon_t = \log(\gamma_t) - \mu_t$ , and let $\mu_t$ define a deterministic function of time containing seasonal and trend components. Then $\varepsilon_{\tau}$ is an OU process:
\begin{align*}
d\varepsilon_{\tau} = -\phi \varepsilon_{\tau}d\tau + \sigma_{\varepsilon} dB_{\tau}
\end{align*}
where $B_\tau$ is standard Brownian motion. It\^{o}'s lemma gives the stochastic differential equation for $\gamma_{\tau} = e^{\varepsilon_{\tau} + \mu_{\tau}}$:
\begin{align*}
d\gamma_\tau = -\phi(\log\gamma_\tau - \mu_\tau)\gamma_\tau + \sigma_{\varepsilon}^2/2\gamma_\tau +d\mu_\tau \gamma_\tau + \sigma_{\varepsilon}\gamma_\tau dB_\tau.
\end{align*}
The solution to the OU process is
\begin{align*}
\varepsilon_{\tau} = e^{-\phi\tau}\left(\varepsilon_0 + \int_0^\tau \sigma_{\varepsilon}e^{\phi s}dB_s\right)
\end{align*}
so that 
\begin{align*}
\varepsilon_{\tau + \delta} | \varepsilon_{\tau} \sim N\left(e^{-\phi\delta}\varepsilon_{\tau}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
\end{align*}
and thus
\begin{align*}
\log \gamma_{\tau + \delta}|\log\gamma_\tau \sim N\left(e^{-\phi\delta}(\log\gamma_{\tau} - \mu_\tau) + \mu_{\tau + \delta}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
\end{align*}
In other words, $\gamma_{\tau + \delta}$ has a lognormal distribution.

To stitch this piece onto the larger model, we exploit the inifnite divisibility property of the gamma distribution. Suppose that
\begin{align*}
x_{\tau + \delta} - x_\tau &\sim G(\beta \int_{\tau}^{\tau + \delta} d\gamma_s, \beta)
\end{align*}
with $d\gamma_\tau$ defined as above. Since both $x_{\tau}$ and $\gamma_{\tau}$ are continuous time processes, it should be possible to break the interval $[\tau, \tau + \delta]$ into a set of subintervals $[\tau + \delta_k, \tau + \delta_{k+1}]$, $0=\delta_0<\delta_1<\dots<\delta_K=\delta$ such that $x_{\tau + \delta} - x_{\tau} = \sum_{k=1}^K (x_{\tau + \delta_k} - x_{\tau + \delta_{k-1}})$. Since $\int_{\tau}^{\tau + \delta}d\gamma_s = \gamma_{\tau + \delta} - \gamma_{\tau}$ we have 
\begin{align*}
x_{\tau + \delta_k} - x_{\tau + \delta_{k-1}} \stackrel{ind}{\sim} G(\beta(\gamma_{\tau + \delta_k} - \gamma_{\tau + \delta_{k-1}}), \beta)
\end{align*}
so that by the gamma distribution's infinite divisibility property
\begin{align*}
\sum_{k=1}^Kx_{\tau + \delta_k} - x_{\tau + \delta_{k-1}} \sim G\left(\beta\sum_{k=1}^K(\gamma_{\tau + \delta_k} - \gamma_{\tau + \delta_{k-1}}), \beta\right) = G\left(\beta(\gamma_{\tau + \delta} - \gamma_\tau), \beta\right).
\end{align*}
In other words, this is consistent with the assumed distribution on $x_{\tau + \delta} - x_{\tau}$.

The full model can then be written as
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G\left(\beta(\gamma_{\tau_k + \delta_k} - \gamma_{\tau_k}),\beta\right)\\
\gamma_{\tau_k + \delta_k}|\gamma_{\tau_k} & \sim LN(\left(e^{-\phi\delta}(\log\gamma_{\tau_k} - \mu_{\tau_k}) + \mu_{\tau_k + \delta_k}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right)\\
\delta_{k} = \tau_{k+1} - \tau_{k} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
for blocks $k=1,2,\dots,N$ where $\mu_{\tau}$ is a parameterized deterministic function of $\tau$. For example, a trend with seasonal variability can be obtained via $\mu_{\tau} = \alpha_0 + \alpha_1\tau + \alpha_2\sin(\omega \tau)$ where $alpha_{0:2}$ is an uknown parameter and $\omega$ is the known frequency of the seasonal effect.

\section{The key question}
As Bitcoin becomes more popular, the transaction data rate increases. Since existing Bitcoin rules limit the number of transactions per block to one MB, at some point the blocks will begin filling up. The purpose of modeling Bitcoin's transaction data rate is to estimate the probability of seeing a full block under a variety of assumptions. Starting from the base model above, the quantity we are interested in is thus
\[
P\left(\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau > 1\right)
\]
for some choice of $k$.

For some distributional forms for the process $dx_\tau$ it may be possible to compute these quantities analytically, but in general it may be difficult. Using a Monte Carlo simulation of $N$ draws, we can approximate these probabilities by
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\int_{\tau_k^{(i)}}^{\tau_{k}^{(i)}+\delta_k^{(i)}}dx_\tau^{(i)} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function. The key parameter we will control in these simulations is the mean transaction data rate, $\gamma$, or the distribution of this rate. Starting with the basic model fit to, say, the last 1000 blocks on the blockchain (approximately one week of elapsed time), we can control this parameter in several ways.

The simplest way to compute parameter estimates based on fitting model, e.g. maximum likelihood estimates, posterior means, or posterior medians. Then estimate the probability of a block filling up by simulating the model conditional on these parameter values. In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\gamma$, while holding other parameters constant.

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases. We can mitigate 1) while making 2) if not worse, different. Specifically, suppose we fit the model and have access to the full (approximate) posterior distribution of the model parameters. Then we can simulate from the posterior of the parameters and for each simulated parameter, simulate the model in order to estimate the probability of a block filling up. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different, e.g. its mean $\bar{\gamma}$. We estimate the probability of a block filling using both approaches, but because of the required assumptions the estimates become less credible for values of $\gamma$ or $\bar{\gamma}$ further away from the estimated values.

Table \ref{tab:probsims} contains estimates of the probability of a filled block under a variety of assumptions, using parameter estimates of the posterior distribution of the model using 1000 contiguous blocks ending with block $253,221$ from August 20, 2013 arriving at about 11:39 a.m.
% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Mon Oct  5 14:56:54 2015
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
$\gamma - \hat{\gamma}$ & $P_{\bar{\gamma}}$ & $P_{\gamma^{(0.5)}}$ & $P_{\mathrm{post}}$ \\ 
  \hline
0 & 0.01 & 0.01 & 0.00 \\ 
  0.01 & 0.03 & 0.03 & 0.03 \\ 
  0.05 & 0.23 & 0.23 & 0.22 \\ 
  0.1 & 0.42 & 0.42 & 0.42 \\ 
  0.5 & 0.82 & 0.82 & 0.83 \\ 
   \hline
\end{tabular}
\caption{Estimated probability of a filled block under a variety of assumptions. From the posterior distribution the mean is $\bar{\gamma} = 0.0179$ and the median is $\gamma^{(0.5)} = 0.0174$. The value of $\gamma - \hat{\gamma}$ is the amount added to the estimated value of $\gamma$ for the purposes of the simulation, $P_{\bar{\gamma}}$ is the probability of a filled block using posterior means for parameter estimates, $P_{\gamma^{(0.5)}}$ is the probability of a filled block using posterior medians, and $P_{\mathrm{post}}$ is the probability of a filled block using the full posterior distribution.}
\label{tab:probsims}
\end{table}

\section{Simulation Exercises}

Each simulation exercise starts with specifying a distribution for $dx_t$. There are three basic categories here: deterministic, i.e. fixed at the mean rate, independent, and temporally dependent. There is actually some interesting room in the deterministic class --- for example, $dx_t$ could be increasing in $t$, or at least not constant. This behavior can be captured in a more realistic stochastic environment, so we will focus on that.

>>>>

\section{Motivation}
As Bitcoin becomes more popular, the transaction rate increases. Since existing Bitcoin rules limit the amount of transactions per block to one MB at some point the blocks will begin filling up. Bitcoin blocks are by design distributed according a Poisson process with a rate of discovery of one block every 10 minutes. Let $\tau$ index time in minutes, and $N(\tau)$ denote the number of blocks discovered by time $\tau$. Then for $\delta > 0$, the number of blocks discovered between time $\tau$ and time $\tau + \delta$ is Poisson distributed, that is $N(\tau + \delta) - N(\tau) \sim Poi(\delta/10)$. Let $k=1,2,\dots$ denote each of the blocks, in order. Then the amount of time between blocks is exponentially distributed. Let $\tau_k$ denote the the time when block $k$ arrives. Then $\delta_k \equiv \tau_{k+1} - \tau_{k} \stackrel{iid}{\sim} Exp(1/10)$ for $k=1,2,\dots,K$, where if $x\sim Exp(\lambda)$ it has the density $p(x) = \lambda e^{-x\lambda}$ for $x>0$ and $\lambda>0$ with mean $1/\lambda$.

There is always some chance that a given block will fill up since arrival times are stochastic. Specifically, for the $k$'th block this probability is
\[
P\left(x_{\tau_k + \delta_k} - x_{\tau_k} > 1\right).
\]
This probability depends crucially on the process that $x_\tau$ follows. If we assume that $x_{\tau}$ is determinstic with a constant transaction rate of $\gamma$ MB/min, then 
\[
x_{\tau_k + \delta_k} - x_{\tau_k} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/10\gamma)
\]
and the probability of a full block is available in closed form as
\[
P\left(x_{\tau_k + \delta_k} - x_{\tau_k} > 1\right) = \int_1^\infty 10\gamma e^{-10\gamma x}dx = e^{-10\gamma}.
\]

Clearly a constant transaction rate does not capture Bitcoin's behavior since it has been growing in popularity. With a non-constant and potentially stochastic transaction rate, the probability of a block filling up is no longer necessarily available in closed form. Using a Monte Carlo simulation of $N$ draws of the block sizes from the process $x_{\tau}$, we can approximate the probability of block $k$ filling up with
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{x^{(i)}_{\tau_{k}^{(i)}+\delta_k^{(i)}} - x^{(i)}_{\tau_k^{(i)}} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function. If we let $\gamma$ denote the mean transaction rate, which is the same as the actual transaction rate in the constant, deterministic case, we can alter $\gamma$ to see how the probability of a full block changes under a variety of regimes.

A key question is whether the assumption of a constant transaction rate is approximately correct on a short enough time-scale. In order to test this assumption we fit a the gamma distribution to the block sizes of various pieces of the block chain. We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(\alpha,\beta)$, it has the density
\[
p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
\]
for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$. Since the $G(1,\beta)$ and $Exp(\beta)$ distributions are identical, if we find that $\alpha\neq 1$ then transaction rates cannot be constant. We focus on the most recent 5000 blocks in our database, which is in the process of downloading the entire chain, ending with block 303151.  For priors we will assume $\log \alpha \sim N(0, 10)$ and $\beta \sim G(1, 1)$. This prior on $\alpha$ implies that the prior median is $\alpha=1$ so that the null hypothesis is favored relatively strongly by the prior. We fit the model to all 5000 blocks, the first 1000 blocks, the second 1000 blocks, etc., and the first 500 blocks, the second 500 blocks, etc. The results are in Table \ref{tab:gammafit}.

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Mon Oct 12 10:44:18 2015
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
 & Mean & 2.5\% & 97.5\% \\ 
  \hline
       $n=5000$ & 1.34 & 1.29 & 1.39 \\ 
$n=1000$, $g=1$ & 1.26 & 1.17 & 1.37 \\ 
$n=1000$, $g=2$ & 1.36 & 1.26 & 1.47 \\ 
$n=1000$, $g=3$ & 1.37 & 1.26 & 1.47 \\ 
$n=1000$, $g=4$ & 1.27 & 1.17 & 1.37 \\ 
$n=1000$, $g=5$ & 1.41 & 1.30 & 1.52 \\ 
 $n=500$, $g=1$ & 1.29 & 1.15 & 1.45 \\ 
 $n=500$, $g=2$ & 1.23 & 1.10 & 1.37 \\ 
 $n=500$, $g=3$ & 1.46 & 1.31 & 1.63 \\ 
 $n=500$, $g=4$ & 1.28 & 1.14 & 1.42 \\ 
 $n=500$, $g=5$ & 1.55 & 1.38 & 1.74 \\ 
 $n=500$, $g=6$ & 1.21 & 1.08 & 1.34 \\ 
 $n=500$, $g=7$ & 1.24 & 1.11 & 1.38 \\ 
 $n=500$, $g=8$ & 1.29 & 1.15 & 1.44 \\ 
 $n=500$, $g=9$ & 1.36 & 1.21 & 1.52 \\ 
 $n=500$, $g=10$ & 1.46 & 1.30 & 1.63 \\ 
 \hline
\end{tabular}
\caption{Posterior mean and 95\% credible intervals for $\alpha$ from fitting a gamma distribution to block sizes of the last 5000 blocks in our database. For sample sizes less than 5000, blocks were broken into groups of equal size and the model was fit separately to each group. Sample size is indicated by $n$, group id by $g$. Each model was fit in RStan\protect\footnotemark using 4 chains, each with 4000 iterations, 2000 of which used for warmup/burn in.}
\label{tab:gammafit}
\end{table}

For\footnotetext{Cite RStan} the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate. This is not surprising since Bitcoin usage has gone up over time, but on a short enough time scale usage may be approximately constant. A block is around 10 minutes on average, so 5000 blocks takes a little over a month on average. It is possible that on a short enough time-scale, constant transaction rates are more appropriate. So we fit the gamma distribution to 1000 blocks at a time --- representing about a week --- by breaking the original 5000 blocks into five groups. For each of these groups the 95\% credible interval does not contain one --- the lowest lower bound is 1.17. Similarly for groups of 500 blocks, each representing about 3.5 days, the lowest lower bound is about 1.08. So we can safely conclude that the transaction rate is not constant. A likely reason for this is seasonality in transaction rates --- Bitcoin trades are probably more likely to happen during certain times of day than others, based on when relevant markets are open and when the bulk of Bitcoin users are awake. The {\it mean} transaction rate averaging over e.g. seasonality is probably approximately constant for time-scales this small, however. In any case, transaction rates are nonconstant even on short time-scales, which motivates modeling them as such.

\section{The Model}

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how how the amount of transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block is completed. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and now $t_k$ the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k} + \delta_k} - x_{\tau_k}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing, infinitely divisible\footnote{Should we define?} stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider the classes of non-negative infinite activity\footnote{Should we define?} L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.\footnote{Cite the Levy process chapter here}. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.

While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error. This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_k + \delta_k}$ without resorting to approximation methods. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments. We will use the gamma process since it is widely available in statistical packages, unlike the main alternative, the inverse Gaussian.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar model that avoids this problem is the lognormal distribution, which we will use.

Combining these pieces we obtain the following model. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
\delta_{k} = \tau_{k+1} - \tau_{k} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ so that block $k$ has $\delta_k$ minutes to accumulate transaction data. To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$. Table \ref{tab:modelfit} contains the results of fitting the model.

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Mon Oct 12 15:04:44 2015
\begin{table}[ht]
  \centering
  \begin{tabular}{lrrrrrrrr}
    \hline
            & Mean    & SD     & SE(Mean) & 2.5\%   & 25\%    & 50\%    & 75\%    & 97.5\% \\ 
    \hline
   $\gamma$ & 0.0240  & 0.0004 & 0.0000   & 0.0232  & 0.0237  & 0.0240  & 0.0243  & 0.0248 \\ 
   $\beta$  & 15.1590 & 0.9432 & 0.0105   & 13.4132 & 14.5042 & 15.1251 & 15.7791 & 17.0910 \\ 
   $\sigma$ & 0.0024  & 0.0001 & 0.0000   & 0.0023  & 0.0024  & 0.0024  & 0.0024  & 0.0025 \\ 
    \hline
 \end{tabular}
 \caption{Results of fitting the measurement error model with stochastic measurement times. We use RStan to fit the model, obtaining four chains each of 4000 iterations, 2000 of which used for warmup/burn in.\protect\footnotemark (Note: outdated numbers, updated ones coming) }
 \label{tab:modelfit}
 \end{table}

%% \subsection{Stochastic transaction rates --- seasonality and autocorrelation}
%% It is unlikely that the transaction rate is constant in time. The transaction rate is clearly growing as Bitcoin becomes more popular, and seasonal variation is expected throughout the data. Further, autocorrelation in transaction rates is likely since periods of high activity will sometimes be caused by, e.g., new information about the long term prospects of Bitcoin.

%% Let $\gamma_\tau$ denote the transaction rate at time $\tau$. We model the evolution of this rate as a variant of the exponential Ornstein–Uhlenbeck (OU) process. Specifically let $\varepsilon_t = \log(\gamma_t) - \mu_t$ , and let $\mu_t$ define a deterministic function of time containing seasonal and trend components. Then $\varepsilon_{\tau}$ is an OU process:
%% \begin{align*}
%% d\varepsilon_{\tau} = -\phi \varepsilon_{\tau}d\tau + \sigma_{\varepsilon} dB_{\tau}
%% \end{align*}
%% where $B_\tau$ is standard Brownian motion. It\^{o}'s lemma gives the stochastic differential equation for $\gamma_{\tau} = e^{\varepsilon_{\tau} + \mu_{\tau}}$:
%% \begin{align*}
%% d\gamma_\tau = -\phi(\log\gamma_\tau - \mu_\tau)\gamma_\tau + \sigma_{\varepsilon}^2/2\gamma_\tau +d\mu_\tau \gamma_\tau + \sigma_{\varepsilon}\gamma_\tau dB_\tau.
%% \end{align*}
%% The solution to the OU process is
%% \begin{align*}
%% \varepsilon_{\tau} = e^{-\phi\tau}\left(\varepsilon_0 + \int_0^\tau \sigma_{\varepsilon}e^{\phi s}dB_s\right)
%% \end{align*}
%% so that 
%% \begin{align*}
%% \varepsilon_{\tau + \delta} | \varepsilon_{\tau} \sim N\left(e^{-\phi\delta}\varepsilon_{\tau}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
%% \end{align*}
%% and thus
%% \begin{align*}
%% \log \gamma_{\tau + \delta}|\log\gamma_\tau \sim N\left(e^{-\phi\delta}(\log\gamma_{\tau} - \mu_\tau) + \mu_{\tau + \delta}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
%% \end{align*}
%% In other words, $\gamma_{\tau + \delta}$ has a lognormal distribution.
